### Workflow for producing a 9 class land cover raster that includes ecosystem condition data (and invasive plant data)

This workflow combines national data on ecosystem condition with invasive plant species and land cover data 2022.

```{r}
#load packages  
library(terra) 
library(tidyverse)
```

```{r}
### Set up data sources & import land cover:

# Base data path from .Renviron
data_path <- Sys.getenv("DATA")

# Locate the required files in the DATA folder and construct a file paths to them

nlc8_fp <- list.files(path = data_path, 
                           pattern = "^lc2022_8class_niaps50\\.tif$",
                           recursive = TRUE,
                           full.names = TRUE)

# Read in the file using terra package
nlc8 <- rast(nlc8_fp)  


# Note  all LC products use  WGS84 Albers Equal Area, central meridian = 25; std parallel1 = -24; std parallel2 = -33). Use Tif version as this is processed faster by R terra

```

```{r}
# Import the Ecosystem Condition Rasters

# Add City of Cape Town Land Cover and Condition 2023 (prepared in ARCGIS PRO), where 0 = inside AOI but no value assigned, 4 = notnat, 2 = sec nat, 8 = severely degraded. 
coct_path <- list.files(
  path = data_path,
  pattern = "^coct_cond10\\.tif$",
  recursive = TRUE,
  full.names = TRUE,
  ignore.case = TRUE
)

coct <- terra::rast(coct_path)
coct <- resample(coct, nlc8, method = "near") #  match extent
#############

# Add NMB Data (prepared in ARCGIS PRO), where 0 = inside AOI but no value assigned, 8 = severely degraded 80%) (ignore value 4 as this is built up areas and national land cover is judged to be a better source of extent data). 
nmb_path <- list.files(
  path = data_path,
  pattern = "^nmb_lcdeg2\\.tif$",
  recursive = TRUE,
  full.names = TRUE,
  ignore.case = TRUE
)

nmb <- terra::rast(nmb_path)
nmb <- resample(nmb, nlc8, method = "near") #  match extent
###############
# Add STEP Data (prepared in ARCGIS PRO), where 0 = inside AOI but no value assigned, 8 = severely degraded 80%). 

step_path <- list.files(
  path = data_path,
  pattern = "^step_deg2\\.tif$",
  recursive = TRUE,
  full.names = TRUE,
  ignore.case = TRUE
)

step <- terra::rast(step_path)
step <- resample(step, nlc8, method = "near") #  match extent
###########
# Add Little Karoo degradation data 2009 (prepared in ARCGIS PRO), where 0 = inside AOI but level of degradation < that 80%; where 8 = severely degraded >= 80%). 
lk_path <- list.files(
  path = data_path,
  pattern = "^littlekaroo_deg4\\.tif$",
  recursive = TRUE,
  full.names = TRUE,
  ignore.case = TRUE
)

lk <- terra::rast(lk_path)
lk <- resample(lk, nlc8, method = "near") #  match extent
###########

# Add Hardeveld degradation data (prepared in ARCGIS PRO), where 0 = inside AOI but level of degradation < that 80%; where 8 = severely degraded >= 80%). 
hv_path <- list.files(
  path = data_path,
  pattern = "^hv_bell\\.tif$",
  recursive = TRUE,
  full.names = TRUE,
  ignore.case = TRUE
)

hv <- terra::rast(hv_path)
hv <- resample(hv, nlc8, method = "near") #  match extent
##########

# Add Little Karoo degradation data (prepared in ARCGIS PRO), where 0 = inside AOI but level of degradation < that 80%; where 8 = severely degraded >= 80%). 

lkkir_path <- list.files(
  path = data_path,
  pattern = "^lk_kirsten\\.tif$",
  recursive = TRUE,
  full.names = TRUE,
  ignore.case = TRUE
)

lkk <- terra::rast(lkkir_path)
lkk <- resample(lk, nlc8, method = "near") #  match extent

```

#### Analysis

Combine the 8 class national land cover with the various ecosystem condition data sets such that IF the 8 class land cover is "natural" (VALUE =1) AND the ecosystem condition for any of the rasters "degraded" (VALUE = 8) THEN pixel is categorised as "degraded" (VALUE = 9) ELSE for pixels values greater than 1 (all other land cover classes) the pixel retains its NLC2022 value.

```{r}
# Combine the rasters such that if nlc8 = 1  AND cond = 8 then value in new raster = 9, else use value of nlc8. This creates a 9 class land cover 1=nat, 2=secnat, 3=artwater, 4=built, 5=crop, 6=mine, 7=plant, 8=invaded (50% cc or more) 9 = severe degradation (overgrazing) 

lc2022_9class_niaps_cond <- ifel(
  (nlc8 == 1) & ((coct == 8) | (step == 8) | (lk == 8) | (lkk == 8)| (hv == 8) | (nmb == 8)),
  9,
  nlc8
)

# Aggregate to 100m resolution 
lc2022_9class_niaps_cond_100m <- aggregate(lc2022_9class_niaps_cond, fact = 5, fun = "modal", na.rm = TRUE)

  # output 100m version 
  # Define subfolder
out_dir <- file.path(data_path, "R_outputs")

# Set output file path and name (out dir previously defined)
outfile1 <- file.path(out_dir, "lc2022_9class_niaps_cond_100m.tif")
# Write a compressed Geo Tiff

writeRaster(lc2022_9class_niaps_cond_100m, outfile1, datatype = "INT1U", gdal = c("COMPRESS=LZW"), overwrite = TRUE)

#output the 20m version 
# Set output file path and name (out dir previously defined)

outfile2 <- file.path(out_dir, "lc2022_9class_niaps_cond.tif")
# Write a compressed Geo Tiff

writeRaster(lc2022_9class_niaps_cond, outfile2, datatype = "INT1U", gdal = c("COMPRESS=LZW"), overwrite = TRUE)
```
